Gender, race, and gun policy during the 2020 US Presidential Election: Biden more favorable among women, Black, Asian, Hispanic, Middle Eastern, and non-gun owners

### 2024-03-16 paper/paper.qmd
Add abstract and discussion section

---
title: "Voter Characteristics and Political Support in the US"
# subtitle: "My subtitle if needed"
author: 
  - Tam Ly
  - Renfrew Ao-Ieong
  - Rahma Binth Mohammad
thanks: "Code and data are available at: https://github.com/atn-ly/political_support_in_the_us"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(rstanarm)
library(knitr)
library(here)
library(ggplot2)
library(arrow)
library(scales)
library(modelsummary)
library(marginaleffects )

@@

Given that Biden and Trump are far apart on women’s issues, racial justice, and gun policy, we  chose *gender*, *race*, and *gunown* to be the predictor variables in our model. First, since we are interested in forecasting who a respondent is likely to vote for, we used *gender* instead of *gender_post* because it is representative of all adult Americans rather than only adult Americans who are registered to vote. Second, we used *race* instead of variables by country or region to affirm that race is a social construct with no biological foundation. Third, we used *gunown* instead of variables about gun control because it captures broader opinions rather than specific policy proposals. After exploring the data in the previous section, we found differences in political preference based on these features and that it would be of interest to investigate further.

@@

To show that our model does a good job of fitting the data, we consider the posterior distribution and implement posterior predictive checks. Details and graphs can be found in @sec-model-details. Furthermore, we check if the Markov Chain Monte Carlo (MCMC) sampling algorithm that `rstanarm` uses to obtain samples from the posterior distributions of interest ran into any issues. Details and graphs can be found in @sec-model-details.

@@

## First discussion point {#sec-first-point}

## Second discussion point

## Third discussion point

@@

>>>>

---
title: "Gender, race, and gun policy during the 2020 US Presidential Election: Biden more favorable among women, Black, Asian, Hispanic, Middle Eastern, and non-gun owners"
# subtitle: "My subtitle if needed"
author: 
  - Tam Ly
  - Renfrew Ao-Ieong
  - Rahma Binth Mohammad
thanks: "Code and data are available at: [https://github.com/atn-ly/political_support_in_the_us](https://github.com/atn-ly/political_support_in_the_us)"
date: today
date-format: long
abstract: "The outcome of the 2020 US Presidential Election served as a reflection of the beliefs and attitudes held by adult Americans. Using data obtained during the fall of 2020 from the Cooperative Election Study, we investigated if a person is likely to vote for Biden, based only on knowing their gender, race, and gun ownership status. The results of our analysis using a Bayesian logistic regression model show that women, Black, Asian, Hispanic, Middle Eastern, and non-gun owners are more likely to support Biden. Our discovery highlights the importance of women’s issues, racial justice, and gun policy on American voters."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

library(tidyverse)
library(rstanarm)
library(knitr)
library(here)
library(ggplot2)
library(arrow)
library(scales)
library(modelsummary)
library(marginaleffects)

@@

Given that Biden and Trump are far apart on women’s issues, racial justice, and gun policy, we  chose “gender”, “race”, and “gunown” to be the predictor variables in our model. First, since we are interested in forecasting who a respondent is likely to vote for, we used “gender” instead of “gender_post” because it is representative of all adult Americans registered to vote rather than only adult Americans who voted. Second, we used “race” instead of variables by country or region to affirm that race is a social construct with no biological foundation. Third, we used “gunown” instead of variables about gun control because it captures broader opinions rather than specific policy proposals. After exploring the data in the previous section, we found differences in political preference based on these features and that it would be of interest to investigate further.

@@

To show that our model does a good job of fitting the data, we consider the posterior distribution and implement posterior predictive checks. Furthermore, we check if the Markov Chain Monte Carlo (MCMC) sampling algorithm that `rstanarm` uses to obtain samples from the posterior distributions of interest ran into any issues. Details and graphs can be found in @sec-model-details.

@@

## Findings

In this paper, data from the 2020 Cooperative Election Study (CES) [@CES] of US political opinion was downloaded and cleaned as discussed in @sec-data. Then, we used a Bayesian logistic regression model as seen in @sec-model to forecast if a person is likely to support Biden, based only on knowing their gender, race, and gun ownership status. Following the outcomes of our analysis in @sec-results, we summarize our three major results here:

1. Women are more likely to vote for Biden,
2. Black, Asian, Hispanic, and Middle Eastern individuals are more likely to vote for Biden, and
3. Non-gun owners are more likely to vote for Biden.

We will now justify and detail the implications of our results by providing the political, historical, cultural, and social context that shaped our story of American political preference in the 2020 US presidential election.

## Women are more likely to vote for Biden

In @sec-results, we found that women have a coefficient of 0.235 and higher estimated probabilities that our model implies of supporting Biden. This means that by just knowing that an individual is female, we can forecast that they are more likely to vote for Biden. Our results imply that Biden’s policies on women’s issues are more favorable than Trump’s. This confirms what we expected as Biden and Trump held opposing stances on various important women’s issues during the 2020 election including women’s reproductive rights, women’s health, and violence against women.

Abortion is a key issue and a deal-breaker for many voters during an election. During Biden’s 2020 campaign, he vowed to protect a women’s right to choose and keep access to abortion legal [@repro_rights]. In particular, Biden promised to protect and codify *Roe v Wade*, a landmark Supreme Court ruling from 1973, which decriminalized abortion nationwide [@repro_rights]. In contrast, Trump pushed for the overturning of *Roe* and appointed conservative anti-abortion judges during his presidency [@repro_rights].

Moreover, women’s health is another important issue that voters consider when deciding which candidate to support. During the 2020 election, Biden supported women’s access to coverage and care [@health]. For example, Biden campaigned to protect the 2010 *Affordable Care Act*, which provides affordable healthcare to millions of women in the US and restricts discriminatory insurance policies [@health]. This contrasts with Trump’s stance on women’s health in which he supported to overturn the entire *ACA* with no plan to replace it [@health].

Furthermore, the issue of sexual violence has been of increasing importance as 1 in 3 women and 1 in 4 men in the US report experiencing sexual violence in their lifetimes [@health]. During his time as a senator, Biden passed the 1994 *Violence Against Women’s Act* which helped establish many violence prevention efforts including rape crisis centers, shelters, and other support services [@health]. During his 2020 campaign, Biden supported reauthorizing and expanding this law, whereas Trump allowed the law to expire in 2018 during his presidency and stated no position on his campaign website [@health].

We learn that Biden and Trump have stark differences on women’s issues and that the outcome of the 2020 election would have major consequences for women. As seen in our findings, women are more likely to vote for Biden, reflecting their support for policies that protect their rights.

## Black, Asian, Hispanic, and Middle Eastern individuals are more likely to vote for Biden

In @sec-results, we found that Black, Asian, Hispanic, and Middle Eastern individuals have coefficients of 2.464, 0.790, 0.548, and 0.288 respectively and higher estimated probabilities that our model implies of supporting Biden. This means that by just knowing that an individual is Black, Asian, Hispanic, or Middle Eastern, we can forecast that they are more likely to vote for Biden. Our results imply that Biden’s policies on racial justice are more favorable within marginalized communities than Trump’s. This aligns with our expectations as Biden and Trump held opposing stances on various important racial issues during the 2020 election including Black Lives Matter, minority-owned small businesses, and housing and education.

In 2020, the police killings of George Floyd in May and Breonna Taylor in March sparked the largest racial justice protests across the US [@BLM]. As a result, racial equality was at the center of many voter’s minds during the 2020 election. Biden’s legacy of Black support was demonstrated by his choice of Kamala Harris as his running mate, as she is the first woman, the first Black American, and the first South Asian American to hold this position [@BLM]. In contrast, Trump called Black Lives Matter “a symbol of hate” and encouraged police violence against protestors [@BLM].

Additionally, Biden and Trump had different strategies to address racial disparities with the US including racial wealth, opportunity, and job gaps [@equality]. In 2020, Biden pledged to support minority-owned small businesses with a 26-page plan by investing $30 billion toward a Small Business Opportunity Fund, $50 billion in venture capital for Black and brown entrepreneurs, and $100 billion toward low-interest loans for new entrepreneurs [@equality]. This is in stark contrast to Trump, who had no broad policy plan to address racial economic inequality [@equality].

Housing and education is another important issue in combating racial inequality in the US. Biden proposed to boost homeownership among minority communities by creating a housing-plan that includes constructing 1.5 million homes and public housing units and ending discriminatory housing policies [@equality]. On education reform, Biden proposed to make public universities and private historically Black colleges and universities tuition-free for families with incomes under $125,000—which the campaign estimates to apply to about 90% of Black, Latino, and Native American households [@equality]. On the other hand, Trump had no plans in place to address these issues again [@equality].

We learn that Biden and Trump are far apart on racial justice and that the 2020 election would shape the lives of many non-white people for years to come . As seen in our findings, Black, Asian, Hispanic, and Middle Eastern individuals are more likely to vote for Biden, reflecting their support for anti-racist policies that promote equality.

## Non-gun owners are more likely to vote for Biden

In @sec-results, we found that an individual with a non-gun owning household has a coefficient of 1.332 and higher estimated probabilities that our model implies of supporting Biden. This means that by just knowing that an individual whose household does not own a gun, we can forecast that they are more likely to vote for Biden. Our results imply that Biden’s policies on gun policies for non-gun owners were more favorable than Trump’s. This confirms what we expected as Biden and Trump held opposing stances on various important policies including gun control, universal background checks, and assault weapons bans.

With the rise in mass shootings and gun-related deaths across the country, gun policies are as important as ever [@guns]. Under the Trump administration, the US saw record-breaking tragedies in the 2017 Las Vegas Strip shooting, 2017 Sutherland Springs church shooting, 2018 Parkland high school shooting, and 2019 El Paso shooting [@guns]. During his presidency, Trump supported the exercise of Second Amendment rights and pushed for greater freedom for gun-owners [@guns2]. He loosened gun regulations on firearms exports and reversed a law that allowed people deemed mentally unfit to purchase guns [@guns2]. In contrast, in Biden’s 2020 campaign, he included plans to require universal background checks for all gun sales and bans on semiautomatic rifles and magazines that contained more than 10 bullets [@guns].

We learn that Biden and Trump have opposite approaches to gun policies and that the outcome of the 2020 election would decide the safety of millions of Americans. As seen in our findings, non-gun owners are more likely to vote for Biden, reflecting their support for policies that prevent gun violence.

@@

Weaknesses arise in the data and model that was used. Although the sampling methodology detailed in @CES is an accepted approach that balances sampling concerns and cost, there are still limitations with large survey data that must be addressed. First, while the survey was not directly conducted by the government, it was conducted by YouGov, an international online research data and analytics technology group, that was funded by the government [@yougov]. This affects our findings since their clients could be Democrats or Republicans that could have their own partisan goals in mind [@yougov]. Historically, there have been cases of data manipulation to suit a government’s narrative in the past [@rohan]. Moreover, there will always be missing data no matter how good the data acquisition process is [@rohan]. CES respondents are people who have made an account on yougov.com or recruited live from online advertisements, and the 2020 CES survey was conducted over the internet [@CES]. However, 1 in 5 US households are not connected to the internet at home [@ntia]. This affects our findings since the sample is representative of Americans with internet access rather than of all national adults as @CES claims.

In future work, we could include additional predictor variables to get a more complete picture of the opinions and traits of voters. For example, the results using family income, educational level, and state of residence as variables, along with their implications, could be of interest. Moreover, we could include the other two presidential candidates of the Libertarian and Green Party in future studies, which requires a different model. Lastly, another aspect to consider could be investigating the post-election questionnaire data. We could build the model again using the post-election data and compare the results with our findings from the pre-election data.
